---
title: RDS_ICA
author: B222908
output:
  word_document
date: "2022-11-10"
---

```{r setup,  include = FALSE}
knitr::opts_chunk$set(echo  =  TRUE)
```

## Import and check data content. Make required genes list unique.

```{r}
# import data
data_all <- read.csv("data_all.csv")
gene_list <- read.csv("genelist_23.txt")
gene_anno <- read.csv("gene_annotation.csv")
sample_anno <- read.csv("sample_annotation.csv")
data_all<-as.data.frame(data_all)

# check if experssion levels in data_all are positive. If not, terminate script
for(i in 1:nrow(data_all)){
  for(k in 1:ncol(data_all)){
    if(data_all[i,k]<0){
      stop("experssion level must be positive number")
      }
  }
}

# check if data_all has duplicated rows and unique it
if (ncol(data_all[duplicated(data_all)]) != 0) {
  data_all<-unique(data_all)
  }
# check if the rows of data_all in order and sort it
if (TRUE %in% (sort(data_all$X) != data_all$X)) {
  data_all<-data_all[order(data_all$X),]
}

# check if gene annotation contain all genes. If not, terminate script
if (FALSE %in% c(data_all$X %in% gene_anno$X)) {
  stop("missing annotation for genes")
}
# check if sample annotation contain all samples. If not, terminate script
if (FALSE %in% c(colnames(data_all)[2:ncol(data_all)] 
                 %in% sample_anno$SampleName)){
  stop("missing annotation for samples")
  }

# remove duplicate genes in required gene list and sort them in order
gene_x <- unique(sort(as.vector(gene_list$x)))
gene_x # print out required genes

```

## Log the expression value and add long name and gene type in the data frame. Print out head of log_data.

```{r}
# log the expression,  remove x column
log_data <- log(data_all)
log_data$X <- data_all$X  # change the X column back to normal
log_data[log_data == -Inf] <- 0  # change the -Inf value into 0

# add new columns of gene types and gene long names in a conservative way
log_data$type=NA
log_data$LongName=NA
for (i in 1:length(log_data$X)){
  # type of gene
  log_data$type[i] <- gene_anno$Type[gene_anno$X==log_data$X[i]]
  # match gene names with long name
  log_data$LongName[i] <- gene_anno$LongName[gene_anno$X==log_data$X[i]]
}
head(log_data)
tail(log_data)
```

## Draw scatter plot.

```{r}

# plot the samples
plot_samples<-function(log_data,gene_x){
# gene types annotation
type_anno <- cbind(gene_x, log_data$type[gene_x])
XA <- type_anno[, 1][type_anno[, 2] == "XA"] # required gene names in XA type
XB <- type_anno[, 1][type_anno[, 2] == "XB"] # required gene names in XB type
XC <- type_anno[, 1][type_anno[, 2] == "XC"] # required gene names in XC type
# define color for each sample in plot. Selected distinguished colors
gene_col <- c("#e6194b",  "#3cb44b",  "#ffe119",  "#4363d8",  "#f58231",
            "#911eb4",  "#46f0f0",  "#f032e6",  "#bcf60c",  "#fabebe",
            "#008080",  "#e6beff",  "#9a6324",  "#fffac8",  "#800000",
            "#aaffc3",  "#808000",  "#ffd8b1",  "#000075",  "#808080")
# plot gene expressions of sample A in type "XA"
plot(XA, log_data[XA, 2], col = gene_col[1], xlab = NA,
     ylab = "log expression", main = "gene expression",
     cex = 0.8, pch = 0, xaxt = "n",
     xlim = c(0, gene_x[length(gene_x)] + 20),
     ylim = c(min(log_data[, 2:(ncol(log_data) - 2)]) - 0.1,
            max(log_data[, 2:(ncol(log_data) - 2)]) + 0.1))
# gene expressions of sample A in other two gene types,
# using different point types(pch)
points(XB, log_data[XB, 2], col = gene_col[1], cex = 0.8, pch = 1)
points(XC, log_data[XC, 2], col = gene_col[1], cex = 0.8, pch = 2)
# long names annotation
axis(side = 1, at = gene_x, labels = log_data$LongName[gene_x],
     cex.axis = 0.7, gap.axis  =  0, las = 3)
# add points of gene expressions from other samples,
# each sample has different color, each gene type has different point type
for (i in 3:(ncol(log_data) - 2)){
  points(XA, log_data[XA, i], col = gene_col[i - 1], cex = 0.8, pch = 0)
  points(XB, log_data[XB, i], col = gene_col[i - 1], cex = 0.8, pch = 1)
  points(XC, log_data[XC, i], col = gene_col[i - 1], cex = 0.8, pch = 2)
}
# add legend of different samples
legend(x = gene_x[length(gene_x)] + 5,
       y = max(log_data[, 2:(ncol(log_data) - 2)]),
       legend  =  colnames(log_data)[2:(ncol(log_data) - 2)],
       col = gene_col[1:(ncol(log_data) - 3)], pch = 1, cex  =  0.8)
# add legend of different gene types
legend(x = 0, y = max(log_data[, 2:(ncol(log_data) - 2)]) + 0.1,
       legend  =  c("XA", "XB", "XC"),
      col = "black", pch = c(0, 1, 2), cex = 0.7, horiz  =  TRUE)
}
plot_samples(log_data,gene_x)
```

## Data processing for drawing heatmap.

```{r}
library(pheatmap)
# create data frame that only contains expression values
exper <- log_data[, 2:(ncol(log_data) - 2)]
row.names(exper) <- log_data$LongName
exper <- exper[gene_x, ] # only required genes
head(exper) # print out the data frame
```

## Annotations for heatmap.

```{r}
# gene type annotation of each gene
gene_type <- data.frame(type = log_data$type) # set a one-column data frame
row.names(gene_type) <- log_data$LongName # row names are gene long names
head(gene_type)
```

```{r}
# treatment annotation of each sample
treatment <- data.frame(treatment = sample_anno$TreatmentGroup)
row.names(treatment) <- sample_anno$SampleName # row names are sample names
head(treatment)
```

## Heatmap 1: both the genes and the samples are clustered. We are interesting in the gene expression between samples, so scale equals to "row".

```{r}
pheatmap(exper, scale = "row", annotation_row  =  gene_type,
         angle_col = "0", annotation_col  =  treatment,
         main = "heatmap of gene expression",
         cluster_rows = TRUE, cluster_cols = TRUE,
         fontsize_row  =  6, fontsize_col  =  12)
```

## Heatmap 2: only the genes are clustered. 

```{r}
pheatmap(exper, scale = "row", annotation_row  =  gene_type,
         angle_col = "0", annotation_col  =  treatment,
         main = "heatmap of gene expression",
         cluster_rows = TRUE, cluster_cols = FALSE,
         fontsize_row = 6, fontsize_col = 12)
```

## Outline the interpretation of the heatmap plots

From the sample clustering, we can see that samples that conducted same treatment show similar gene expression patterns. In addition, the expression patterns of treatment 2 and 3 show great similarities, while expression pattern of treatment 4 is significantly different from expression patterns of other groups. 

From the gene type clustering, we can see that genes that are in the same type don't cluster together, which reveals that the expression patterns of genes in same type doesn't necessarily to be similar. Moreover, we can see that some genes shows similar changes across different treatments. The expression of genes significantly changed between treatments can become biomarker, showing the biological changes when those treatments are conducted.


